---
title: "NHL NO GIFT"
output: html_document
date: "2024-12-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
---

### **Purpose**
- The Purpose of this analysis is to determine the historical hit rate of a dummy leg where 0 goals are scored in the first 10 minutes of an NHL hockey game and 3 or more are scored in the second 10 minutes.

---

---

### **1. Library Setup**
- **Overview**:
  - Loads the `readr` and `dplyr` libraries, which are essential for reading data files and manipulating data frames efficiently.
- **Purpose**:
  - Prepares tools for handling data input, cleaning, and summarizing.

---

```{r, message = FALSE}
#load libraries
library(readr)
library(dplyr)
```
### **2. Data Acquisition**
- **Overview**:
  - Downloads a compressed dataset of NHL shot data (2015–2023) from a given URL. It extracts the data file and reads it into an R data frame for analysis.
- **Purpose**:
  - Ensures access to raw shot data directly from the source and loads it into a usable format.

---


```{r, message = FALSE}
# shot data gathered from https://moneypuck.com/data.htm
# load Recent Seasons (2015-2023 Seasons) (1,006,503 shots)
# Set the URL of the zipped CSV file
url <- "https://peter-tanner.com/moneypuck/downloads/shots_2015-2023.zip"

# Set the path for the temporary file
temp_file_path <- tempfile(fileext = ".zip")

# Download the file from the URL and save it to the temporary file
download.file(url, destfile = temp_file_path, mode = "wb")

# Unzip the file to the temporary directory
temp_dir <- tempdir()
unzip(temp_file_path, exdir = temp_dir)
filename <- unzip(temp_file_path, exdir = temp_dir, list = TRUE)$Name[1]

#read file into workspace
shot_data <- readr::read_csv(file.path(temp_dir,filename))
```


### **3. Initial Data Filtering**
- **Overview**:
  - Selects only the relevant columns from the dataset (shot ID, season, game ID, time, goal indicator, and period). Filters data to focus on events occurring during the first period of games.
- **Purpose**:
  - Reduces the dataset to focus on specific events that are crucial for the analysis, such as period-specific goals.

---

```{r}
#subset to columns we're interested in
shot_time_game <- shot_data |> select(shotID,season,game_id,time,goal,period)

# only care about goals
first_period_shots <- shot_time_game |> filter(period == 1)
```

### **4. Goal Summarization**
- **Overview**:
  - Groups data by season and game ID, then calculates the total number of goals scored during two specific timeframes: the first 10 minutes and the second 10 minutes of the first period.
- **Purpose**:
  - Provides key metrics for evaluating scoring trends in different parts of the game.

---
```{r}
# encode number of goals in first 10 versus second 10
first_ten_goals <- first_period_shots |> filter(time<=600) |> group_by(season,game_id) |> summarise(first_ten_goals = sum(goal))
second_ten_goals <- first_period_shots |> filter(time>600) |> group_by(season,game_id) |> summarise(second_ten_goals = sum(goal))
```



### **5. Combining Summaries**
- **Overview**:
  - Merges the datasets containing the number of goals scored in the first 10 minutes and the second 10 minutes into a single summary table.
- **Purpose**:
  - Consolidates the data to allow comparison and further analysis of goals scored in the two timeframes.

---
```{r}
# join together
first_period_goals <- first_ten_goals |> left_join(second_ten_goals)
```



### **6. Filtering Specific Game Criteria**
- **Overview**:
  - Identifies games that meet specific conditions, such as having no goals in the first 10 minutes but three or more goals in the second 10 minutes.
- **Purpose**:
  - Pinpoints rare or notable scoring patterns for deeper exploration.

---

```{r}
# find games where there was no goals in first 10 but 3 or more in second 10
dummy_leg_calc <- first_period_goals |> filter(first_ten_goals == 0 & second_ten_goals >= 3)
```

### **7. Link Generation**
- **Overview**:
  - Creates clickable URLs for each game that meets the filtering criteria, linking to detailed game data hosted on an external website.
- **Purpose**:
  - Enables quick access to additional resources or visualizations for further analysis.

---

```{r}
#format example urls
dummy_leg_calc_url <- dummy_leg_calc |> rowwise() |> mutate(url = paste0("<a href='https://moneypuck.com/g.htm?id=", season, 0, game_id,"'>",game_id,"</a>",collapse = ""))
#show examples
library(DT)
datatable(dummy_leg_calc_url |> arrange(desc(second_ten_goals)) |> select(-game_id),escape = FALSE)
```

### **8. Percentage Calculation**
- **Overview**:
  - Calculates the percentage of games where the specific conditions (e.g., three or more second-half goals with no first-half goals) did not occur.
- **Purpose**:
  - Summarizes the rarity or frequency of the analyzed pattern in the dataset.

---


```{r}
# calculate hit percent
100 - 100*nrow(dummy_leg_calc)/nrow(first_period_goals)
```

